Report on Late Winter2013 Production: Image Differencing\vspace{-2ex}

### 2013-04-01 W13report.tex
Fixed footnote symbols

\renewcommand{\thefootnote}{\fnsymbol{footnote}}

@@

We investigated the complexity of the basis functions used to model
the PSF--matching kernel, and the degree to which this kernel varies
across the images.  We generate several goodness of fit statistics for
a control sample of sources that were {\it not} used in the kernel fit
(XXX name these metrics), in order to assess the interpolation ability
of the matching kernel.  These metrics are shown to correlate with the
numbers of false positives.  We calculate a mean squared error metric
that shows a minimum value close to the configurations that yielded
the lowest number of false positives, and shows promise as a means to
algorithmically choose the optimal pipeline configuration.

@@

We outline the image subtraction process below.  We assume that science
image $S(x,y)$ can be modeled as a convolution of the template image
$T(x,y)$ by a PSF--matching kernel $K(u,v,x,y)$.  The images will have
different point spread functions (PSFs), which are the time--averaged
transfer functions of a point source through the Earth's atmosphere,
telescope optics, and into the silicon of the detector before being
read out.  The essence of image subtraction is to match the PSFs of
these two images so that they may be subtracted pixel by pixel.  We
further assume that the PSF--matching kernel may be decomposed using a
set of basis functions $K(u,v) = \sum_i a_i K_i(u,v)$, where the
coefficients in front of each basis are determined through:
%
\begin{eqnarray}
C_i & \equiv & (K_i \otimes T) \\ \nonumber
b_{i}  & = & \sum_{x,y} {{C_i(x,y) S(x,y)}\over{\sigma^2(x,y)}}   \nonumber \\ 
M_{ij} & = & \sum_{x,y} {{C_i(x,y) C_j(x,y)}\over{\sigma^2(x,y)}}  \nonumber \\ 
a_{i}  & = & M^{-1}_{ij} b_{j} \nonumber 
\label{eq-soln}
\end{eqnarray}
where $\sigma^2(x,y)$ is the per--pixel variance stored in the {\tt
  variance} plane of each LSST {\tt exposure}.  To generate a
spatially varying model for the kernel, we assume that the relative
weights of the basis functions $a_i$ themselves vary spatially,
i.e. $K(u,v,x,y) = \sum_i a_i(x,y) K_i(u,v)$.  We also assume that
there is a differential background between the two images $b(x,y)$
that may be fit for using a low--order Chebyshev polynomial.  The
image difference is then calculated through $D(x,y) = S(x,y) - T(x,y)
\otimes K(u,v,x,y) - b(x,y)$.

@@

Detection on the difference image occurs through correlation of
$D(x,y)$ with the science image's PSF, yielding optimally filtered
detection image $D'(x,y) = D(x,y) \circ PSF_S(u,v,x,y)$ where $\circ$
denotes correlation.  The values of the pixels in $D'(x,y)$ provide a
maximum likelihood estimate of there being a point source at that
position.  Detection occurs by simply finding pixels that are more
than N sigma above the square root of the per--pixel variance.

In this production, we investigated two orders of operations for the
optimal filtering by the PSF.  In the first method, called
prefiltering, we PSF--match the template image to a pre--filtered
science image:
\begin{eqnarray}
D_{Pre}(x,y) & = & S(x,y) \circ PSF_S(u,v,x,y) - T(x,y) \otimes K(u,v,x,y) \nonumber \\ 
D'_{Pre}(x,y) & = & D_{Pre}(x,y). \nonumber 
\end{eqnarray}
By pre--smoothing the image that we are matching the template to --
$S(x,y) \circ PSF_S(u,v,x,y)$ -- we more frequently avoid the case of
PSF--matching {\it deconvolution}, where the FWHM of $T$ is broader
than $S$.  This pre--filtering increases $\sigma_S$ by $\sqrt{2}$.
The PSF matching kernels will need to be correspondingly larger to
account for the larger effective PSFs.  We are able to run detection
directly on this image.
%XXX K(x,y,mu,nu) is pretty confusing  - especially as it has an
%explicit sum in it for a number of the PSF detection steps (ie sum
%over mu,nu)

In the second, which is the classical postfiltering implementation of
the technique, we create a difference image that is then
post--filtered with the PSF for detection:
\begin{eqnarray}
D_{Post}(x,y) & = & S(x,y) - T(x,y) \otimes K(u,v,x,y) \nonumber \\ 
D'_{Post}(x,y) & = & D_{Post}(x,y) \circ PSF_S(u,v,x,y).  \nonumber 
\end{eqnarray}

@@

Next, we used the DiaCatalogSourceSelector to query the ImSim
reference catalog for appropriate sources to use for PSF matching.
This selector allows the user to specify the brightness and color
range of the objects, toggle star or galaxy selection, and to include
variable objects or not.  We selected a fraction (1/5) of these
objects to serve as a control sample, in order to assess the
effectiveness of $K(u,v,x,y)$ to subtract objects that were {\it not}
used in the least squares fit.

@@

Images $S(x,y)$ (optionally pre-filtered with $PSF_S(u,v,x,y)$) and
$T(x,y)$ were sent to ImagePsfMatchTask, whose purpose is to fit for
$K(u,v,x,y)$ and to produce the difference image $D(x,y)$.

@@

 The first step is to use the
PSF Gaussian sigmas to define the {\it sizes} of the Gaussians shapes
in $K_i(u,v)$.  As mentioned above, the main width was chosen to be
$\sqrt{\sigma_S^2 - \sigma_T^2}$.  We investigated variations on the
configuration where we used one Gaussian with the above width, and
where we used three Gaussians, the central of which has this width and
the others being a factor of $\beta = 2.0$ smaller and larger.  We
modified each of the Gaussians by a set of Laguerre polynomials of a
given order.  The smallest Gaussian was modified to the specified
order, with the others modified by order//2 {XXX what is this?}.  Figure~\ref{basis}
provides an example basis set with this configuration.  The total
number of shapes in each basis were: $\sum_i^{\rm nGauss} ({\rm
  degGauss}_i+1)\times({\rm degGauss}_i+2)/2$ \footnote{ A notable
  exception to this algorithm is found in the case where $\sigma_T >
  \sigma_S$, i.e. a sharpening must be done to the template to match
  it to the science image.  This is the case of deconvolution.  A
  theoretical prescription for basis design in this case is given in
  \cite{0266-5611-26-8-085002}.  In our implementation of this algorithm, nGauss is {\it
    always} fixed to value 3, and degGauss is always fixed to 3.  The
  widths of the Gaussians are determined through the algorithm
  specified in \cite{0266-5611-26-8-085002}, using as inputs the
  sequence of Gaussians that we would have used to match a Gaussian of
  width $\sigma_S$ to $\sigma_T$ (i.e. as if we would have convolved
  the science image and not the template image).  There are a variety
  of reasons that actually operating on the science image is
  undesirable, including: the science image is lower S/N than the
  template; the science image is likely to have masked defects,
  including cosmic rays and bad pixels, whose reach would be expanded
  due to the convolution; unmasked features such as faint cosmic rays
  are more likely to look PSF--like after convolution (cosmic rays
  will likely have been sigma clipped out of the coadd).  However, it
  is not clear that deconvolving the template image is in principle
  any better.  A far better option, which we explore here, is to
  pre--filter the science image, which will in the vast majority of
  cases provide a larger PSF (by a factor of $\sqrt{2}$) to match the
  template to.  This is the case for v6866601 in this study, where the
  postfiltered analysis yields deconvolution of the template, while
  the prefiltered analysis does not.  }.
%{XXX footnote should note be a * as it confuses the math...
The dimensions of the PSF matching kernel were chosen to be 6 times
the largest Gaussian width, with a minimum kernelSize of 21x21 pixels.
For kernels that are significantly smaller than this, the Gaussians
have significant (non--zero) power at the kernel boundaries, leading
to square systematic artifacts at the scale of the kernel in the
difference images (see Figure~\ref{fig_galleryBad}).

@@

For each source $j$ returned by DiaCatalogSourceSelector, we created a
{\tt KernelCandidate} instance, which holds substamps of $T_j(x,y)$
and $S_j(x,y)$ centered on the source.  These were used to fit for a
local solution $K_j(u,v)$; the ensemble of local solutions at each
source's $x,y$ was used to fit for the full spatial solution
$K(u,v,x,y)$.  We emphasize that the dimensions of the substamps are
important.  Note from equation~\ref{eq-soln} that the coefficients
$a_i$ are derived from a convolution of the template substamp with the
kernel basis functions $K_i$.  This convolution means that
kernelSize//2 pixels along the border of each substamp are rendered unusable.
To maintain a significant number of pixels in $C_i$, we set the {\tt
  KernelCandidate} dimensions to be twice that of the kernel (i.e. the
number of pixels remaining in $C_i$ is equal to the number of pixels
in the kernel, which is fewer than the number of bases $i$).  For each
{\tt KernelCandidate}, we solved for $K_j(u,v)$ and create a local
difference image $D_j(x,y)$.  We evaluated several statistics on this
local difference image, normalized by the square root of its variance,
which puts the pixels in units of sigma (see also
Section~\ref{sec-stats}).  We measured the mean sigma, the RMS of this
distribution, the $\chi^2$ of these pixels normalized by the degrees
of freedom (number of pixels in $D_j(x,y)$ minus the number of kernel
bases), and the mean squared error.  These are referred to as the
LOCAL kernel metrics, and reflect how appropriate the chosen kernel
basis is (see again Figure~\ref{fig_galleryBad}).

The ensemble of {\tt KernelCandidates} was then used to constrain a
spatial model of the kernel $K(u,v,x,y)$.  This was done using the
{\tt SpatialCellSet} formalism, whereby the {\tt KernelCandidates} are
distributed across the image in a grid of {\tt SpatialCells}.  This
formalism attempts to distribute constraints on the spatial model
evenly across the image.  At most 3 {\tt KernelCandidates} per {\tt
  SpatialCell} were input to the spatial kernel fit.  For the spatial
kernel model, we assumed that each of the kernel coefficients $a_i$
may be represented by a $N^{th}$ order 2--dimensional Chebyshev.
Normally, we would perform sigma--clipping iterations on candidates
having bad residuals after spatial modeling.  However in this
production we expect that all objects should fit, so we performed no
sigma clipping.  Having generated the full spatial solution
$K(u,v,x,y) = \sum_i a_i(x,y) K_i(u,v)$, we evaluated the kernel
solution at the position of each {\tt KernelCandidate}, created a
difference image using this kernel, and recalculated the metrics
defined above.  These are referred to as the SPATIAL kernel metrics.
Importantly, we also interpolated this solution to the positions of
the control sample, which were explicitly {\it not} used in the
spatial fit, and evaluate these same metrics.  The differences between
the {\tt KernelCandidate} LOCAL and SPATIAL metrics, as well as the
values of the SPATIAL metrics for the control sample, reflect the
appropriateness of the spatial model and our ability to interpolate
and extrapolate to the full extent of the images.  We used the Mean
Square Error (MSE) of the control sample to example the tradeoff
between bias and variance in the spatial interpolation.

@@

We implemented several diagnostics on the quality of the image
subtractions for a given {\tt KernelCandidate}.  This was done for the
original $T_j(x,y)$ to $S_j(x,y)$ kernel fit, yielding LOCAL quality
statistics.  These were re--evaluated using the spatial kernel model
evaluated at the {\tt KernelCandidate} location $K(u,v,x_j,y_j)$ ,
yielding SPATIAL quality statistics.  The task metadata was augmented
with summary statistics of these values (mean, median, RMS), and the
information on each {\tt KernelCandidate} persisted in a new {\tt
  kernelSrc} data product.  We briefly describe more useful of these
statistics below.

>>>>

%\renewcommand{\thefootnote}{\fnsymbol{footnote}}

\makeatletter
\newcommand*{\myfnsymbolsingle}[1]{%
  \ensuremath{%
    \ifcase#1% 0
    \or % 1
      \mathparagraph%   
    \or % 2
      \dagger
    \or % 3  
      \ddagger
    \or % 4   
      \mathsection
    \or % 5
      \mathparagraph
    \else % >= 6
      \@ctrerr  
    \fi
  }%   
}   
\makeatother
\newcommand*{\myfnsymbol}[1]{%
  \myfnsymbolsingle{\value{#1}}%
}
\renewcommand*{\thefootnote}{%
  \myfnsymbolsingle{\value{footnote}}%
}

@@

We investigated the complexity of the basis functions used to model
the PSF--matching kernel, and the degree to which this kernel varies
across the images.  We generate several goodness of fit statistics for
a control sample of sources that were {\it not} used in the kernel fit
notably the mean squared error, in order to assess the interpolation ability
of the matching kernel.  These metrics are shown to correlate with the
numbers of false positives.  We calculate a mean squared error metric
that shows a minimum value close to the configurations that yielded
the lowest number of false positives, and shows promise as a means to
algorithmically choose the optimal pipeline configuration.

@@

We outline the image subtraction process below.  We assume that science
image $S(x,y)$ can be modeled as a convolution of the template image
$T(x,y)$ by a PSF--matching kernel $K(u,v;x,y)$ (indices $u,v$ indicate
that the kernel itself is a 2--dimensional function, which varies as a
function of position $x,y$ in the image.  During convolution and correlation
there is an implicit summation over $u,v$).  The images will have
different point spread functions (PSFs), which are the time--averaged
transfer functions of a point source through the Earth's atmosphere,
telescope optics, and into the silicon of the detector before being
read out.  The essence of image subtraction is to match the PSFs of
these two images so that they may be subtracted pixel by pixel.  We
further assume that the PSF--matching kernel may be decomposed using a
set of basis functions $K(u,v) = \sum_i a_i K_i(u,v)$, where the
coefficients in front of each basis are determined through:
%
\begin{eqnarray}
C_i & \equiv & (K_i \otimes T) \\ \nonumber
b_{i}  & = & \sum_{x,y} {{C_i(x,y) S(x,y)}\over{\sigma^2(x,y)}}   \nonumber \\ 
M_{ij} & = & \sum_{x,y} {{C_i(x,y) C_j(x,y)}\over{\sigma^2(x,y)}}  \nonumber \\ 
a_{i}  & = & M^{-1}_{ij} b_{j} \nonumber 
\label{eq-soln}
\end{eqnarray}
where $\sigma^2(x,y)$ is the per--pixel variance stored in the {\tt
  variance} plane of each LSST {\tt exposure}.  To generate a
spatially varying model for the kernel, we assume that the relative
weights of the basis functions $a_i$ themselves vary spatially,
i.e. $K(u,v;x,y) = \sum_i a_i(x,y) K_i(u,v)$.  We also assume that
there is a differential background between the two images $b(x,y)$
that may be fit for using a low--order Chebyshev polynomial.  The
image difference is then calculated through $D(x,y) = S(x,y) - T(x,y)
\otimes K(u,v;x,y) - b(x,y)$.

@@

Detection on the difference image occurs through correlation of
$D(x,y)$ with the science image's PSF, yielding optimally filtered
detection image $D'(x,y) = D(x,y) \circ PSF_S(u,v;x,y)$ where $\circ$
denotes correlation.  The values of the pixels in $D'(x,y)$ provide a
maximum likelihood estimate of there being a point source at that
position.  Detection occurs by simply finding pixels that are more
than N sigma above the square root of the per--pixel variance.

In this production, we investigated two orders of operations for the
optimal filtering by the PSF.  In the first method, called
prefiltering, we PSF--match the template image to a pre--filtered
science image:
\begin{eqnarray}
D_{Pre}(x,y) & = & S(x,y) \circ PSF_S(u,v;x,y) - T(x,y) \otimes K(u,v;x,y) \nonumber \\ 
D'_{Pre}(x,y) & = & D_{Pre}(x,y). \nonumber 
\end{eqnarray}
By pre--smoothing the image that we are matching the template to --
$S(x,y) \circ PSF_S(u,v;x,y)$ -- we more frequently avoid the case of
PSF--matching {\it deconvolution}, where the FWHM of $T$ is broader
than $S$.  This pre--filtering increases $\sigma_S$ by $\sqrt{2}$.
The PSF matching kernels will need to be correspondingly larger to
account for the larger effective PSFs.  We are able to run detection
directly on this image.

In the second, which is the classical postfiltering implementation of
the technique, we create a difference image that is then
post--filtered with the PSF for detection:
\begin{eqnarray}
D_{Post}(x,y) & = & S(x,y) - T(x,y) \otimes K(u,v;x,y) \nonumber \\ 
D'_{Post}(x,y) & = & D_{Post}(x,y) \circ PSF_S(u,v;x,y).  \nonumber 
\end{eqnarray}

@@

Next, we used the DiaCatalogSourceSelector to query the ImSim
reference catalog for appropriate sources to use for PSF matching.
This selector allows the user to specify the brightness and color
range of the objects, toggle star or galaxy selection, and to include
variable objects or not.  We selected a fraction (1/5) of these
objects to serve as a control sample, in order to assess the
effectiveness of $K(u,v;x,y)$ to subtract objects that were {\it not}
used in the least squares fit.

@@

Images $S(x,y)$ (optionally pre-filtered with $PSF_S(u,v;x,y)$) and
$T(x,y)$ were sent to ImagePsfMatchTask, whose purpose is to fit for
$K(u,v;x,y)$ and to produce the difference image $D(x,y)$.

@@

 The first step is to use the
PSF Gaussian sigmas to define the {\it sizes} of the Gaussians shapes
in $K_i(u,v)$.  As mentioned above, the main width was chosen to be
$\sqrt{\sigma_S^2 - \sigma_T^2}$.  We investigated variations on the
configuration where we used one Gaussian with the above width, and
where we used three Gaussians, the central of which has this width and
the others being a factor of $\beta = 2.0$ smaller and larger.  We
modified each of the Gaussians by a set of Laguerre polynomials of a
given order.  The smallest Gaussian was modified to the specified
order, with the others modified by order/2.  Figure~\ref{basis}
provides an example basis set with this configuration.  The total
number of shapes in each basis were: $\sum_i^{\rm nGauss} ({\rm
  degGauss}_i+1)\times({\rm degGauss}_i+2)/2$ \footnote{ A notable
  exception to this algorithm is found in the case where $\sigma_T >
  \sigma_S$, i.e. a sharpening must be done to the template to match
  it to the science image.  This is the case of deconvolution.  A
  theoretical prescription for basis design in this case is given in
  \cite{0266-5611-26-8-085002}.  In our implementation of this algorithm, nGauss is {\it
    always} fixed to value 3, and degGauss is always fixed to 3.  The
  widths of the Gaussians are determined through the algorithm
  specified in \cite{0266-5611-26-8-085002}, using as inputs the
  sequence of Gaussians that we would have used to match a Gaussian of
  width $\sigma_S$ to $\sigma_T$ (i.e. as if we would have convolved
  the science image and not the template image).  There are a variety
  of reasons that actually operating on the science image is
  undesirable, including: the science image is lower S/N than the
  template; the science image is likely to have masked defects,
  including cosmic rays and bad pixels, whose reach would be expanded
  due to the convolution; unmasked features such as faint cosmic rays
  are more likely to look PSF--like after convolution (cosmic rays
  will likely have been sigma clipped out of the coadd).  However, it
  is not clear that deconvolving the template image is in principle
  any better.  A far better option, which we explore here, is to
  pre--filter the science image, which will in the vast majority of
  cases provide a larger PSF (by a factor of $\sqrt{2}$) to match the
  template to.  This is the case for v6866601 in this study, where the
  postfiltered analysis yields deconvolution of the template, while
  the prefiltered analysis does not.  }.
%{XXX footnote should note be a * as it confuses the math...
The dimensions of the PSF matching kernel were chosen to be 6 times
the largest Gaussian width, with a minimum kernelSize of 21x21 pixels.
For kernels that are significantly smaller than this, the Gaussians
have significant (non--zero) power at the kernel boundaries, leading
to square systematic artifacts at the scale of the kernel in the
difference images (see Figure~\ref{fig_galleryBad}).

@@

For each source $j$ returned by DiaCatalogSourceSelector, we created a
{\tt KernelCandidate} instance, which holds substamps of $T_j(x,y)$
and $S_j(x,y)$ centered on the source.  These were used to fit for a
local solution $K_j(u,v)$; the ensemble of local solutions at each
source's $x,y$ was used to fit for the full spatial solution
$K(u,v;x,y)$.  We emphasize that the dimensions of the substamps are
important.  Note from equation~\ref{eq-soln} that the coefficients
$a_i$ are derived from a convolution of the template substamp with the
kernel basis functions $K_i$.  This convolution means that
kernelSize//2 pixels along the border of each substamp are rendered unusable.
To maintain a significant number of pixels in $C_i$, we set the {\tt
  KernelCandidate} dimensions to be twice that of the kernel (i.e. the
number of pixels remaining in $C_i$ is equal to the number of pixels
in the kernel, which is fewer than the number of bases $i$).  For each
{\tt KernelCandidate}, we solved for $K_j(u,v)$ and create a local
difference image $D_j(x,y)$.  We evaluated several statistics on this
local difference image, normalized by the square root of its variance,
which puts the pixels in units of sigma (see also
Section~\ref{sec-stats}).  We measured the mean sigma, the RMS of this
distribution, the $\chi^2$ of these pixels normalized by the degrees
of freedom (number of pixels in $D_j(x,y)$ minus the number of kernel
bases), and the mean squared error.  These are referred to as the
LOCAL kernel metrics, and reflect how appropriate the chosen kernel
basis is (see again Figure~\ref{fig_galleryBad}).

The ensemble of {\tt KernelCandidates} was then used to constrain a
spatial model of the kernel $K(u,v;x,y)$.  This was done using the
{\tt SpatialCellSet} formalism, whereby the {\tt KernelCandidates} are
distributed across the image in a grid of {\tt SpatialCells}.  This
formalism attempts to distribute constraints on the spatial model
evenly across the image.  At most 3 {\tt KernelCandidates} per {\tt
  SpatialCell} were input to the spatial kernel fit.  For the spatial
kernel model, we assumed that each of the kernel coefficients $a_i$
may be represented by a $N^{th}$ order 2--dimensional Chebyshev.
Normally, we would perform sigma--clipping iterations on candidates
having bad residuals after spatial modeling.  However in this
production we expect that all objects should fit, so we performed no
sigma clipping.  Having generated the full spatial solution
$K(u,v;x,y) = \sum_i a_i(x,y) K_i(u,v)$, we evaluated the kernel
solution at the position of each {\tt KernelCandidate}, created a
difference image using this kernel, and recalculated the metrics
defined above.  These are referred to as the SPATIAL kernel metrics.
Importantly, we also interpolated this solution to the positions of
the control sample, which were explicitly {\it not} used in the
spatial fit, and evaluate these same metrics.  The differences between
the {\tt KernelCandidate} LOCAL and SPATIAL metrics, as well as the
values of the SPATIAL metrics for the control sample, reflect the
appropriateness of the spatial model and our ability to interpolate
and extrapolate to the full extent of the images.  We used the Mean
Square Error (MSE) of the control sample to example the tradeoff
between bias and variance in the spatial interpolation.

@@

We implemented several diagnostics on the quality of the image
subtractions for a given {\tt KernelCandidate}.  This was done for the
original $T_j(x,y)$ to $S_j(x,y)$ kernel fit, yielding LOCAL quality
statistics.  These were re--evaluated using the spatial kernel model
evaluated at the {\tt KernelCandidate} location $K(u,v;x_j,y_j)$ ,
yielding SPATIAL quality statistics.  The task metadata was augmented
with summary statistics of these values (mean, median, RMS), and the
information on each {\tt KernelCandidate} persisted in a new {\tt
  kernelSrc} data product.  We briefly describe more useful of these
statistics below.
